Investigating motor vehicle collisions in Toronto: Installing more red light cameras do not stop traffic accidents from happening

### 2022-02-05 outputs/paper/paper.Rmd
finish data

---
title: "Toronto motor vehicle collisions can be reduced by installing more red cameras"
subtitle: "Toronto motor vehicle collisions can be reduced by installing more red cameras"
author: 
  - Alicia Yang
thanks: "Code and data are available at: https://github.com/Alicia-y/Telling-stories-with-data."
date: "`r format(Sys.time(), '%d %B %Y')`"
abstract: "Vehicle collisions occur constantly in big citys. They sometimes could be severe issues when resulting in damage in public and personal properties, injuries, or even death. This report aims to investigate the motor vehicle collisions in Toronto that result in injuries or deaths from 2006 to 2020. By exploring the trend of these traffic collisions in respect of number and location, I find that .... Implies that ... The data used in this report is obtained from the City of Toronto Open Data Portal, and ."
output:
  bookdown::pdf_document2
toc: TRUE
bibliography: references.bib
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)

library(tidyverse)
library(janitor)
library(palmerpenguins)
```

@@

As one of the busiest and most modern city in Canada, Toronto constantly faces a large number of citizens, and, inevitably, heavy traffic. The road condition in Toronto is busy most of the time. The vast majority of roads and buildings under construction that take place almost everywhere narrow the roadway of vehicles and turn off traffic lights. Heavy traffic is likely to be one of the main reasons behind car accidents, because it increases the road rage and encourages the vehicle drivers to make dangerous moves to save time. Vehicle collisions have always brought large financial costs to the involving parties, and even more, they even brought injuries and deaths.

The Motor Vehicle Collisions data of Toronto provides a record of the past severe vehicle collisions in Toronto which result in injuries or deaths, and with the help of statistical data analysis, it can possibly offer insights on the ways to reduce this type of accident. This report investigates the motor vehicle collisions in Toronto from 2006 to 2020, with a focus on the ones result in injuries or deaths. In Section \@ref(data), the Motor Vehicle Collisions dataset is introduced and organized to visualize the change in the annual number of accidents, as well as comparing the frequency of the accidents by district, accident type, and level of injury. Then, I discussed about the possible implications of these visualizations.

However, numbers of annual motor vehicle collisions demonstrate a generally descending trend from 2006 to 2020.

@@

# Data
To investigate the motor vehicle collisions in Toronto, I obtained the dataset "Motor Vehicle Collisions involving Killed or Seriously Injured Persons" from the Toronto Open Data Portal. Provided by the Toronto Police Service, this dataset includes all vehicle collisions occurred in Toronto between 2006 and 2020 that result in injuries or deaths. It contains 16840 observations of collisions and 54 variables with all kinds of information of each collision. This report focuses on six of these variables which are: year, accident location, district, street1(of the collision occurred), street2(of the collision occurred), and classification of accident. By using R packages "tidyverse" [@citetidyverse], "janitor" [@citejanitor], "kableExtra" [@citekableExtra] and "knitr" (Yihui Xie 2021), an extract of the cleaned dataset (Table \@ref(tab:dataextract)) is shown below.

```{r, echo=FALSE, message=FALSE, warning=FALSE}
vehicle_collision <- read_csv(here::here("Telling-stories-with-data/inputs/data/vehicle_collisions.csv")) %>%
  clean_names() %>%
  drop_na(year)
```

```{r dataextract, echo=FALSE, message=FALSE}
library(knitr)
library(kableExtra)
vehicle_collision <- vehicle_collision %>%
  select(year, street1, street2, accloc, district, acclass)
vehicle_collision %>%
  slice(1:10) %>%
  kable(
    caption = "Extracting the first ten rows from the Motor Vehicle Collisions data",
col.names=c("Year", "Street1", "Street2", "Collision location", "District", "Classification of Accident

"),
linesep = "", digits = 1, booktabs=TRUE) %>%
  kable_styling(latex_options = "HOLD_position")

@@

Table \@ref(tab:dataextract) shows the first ten rows of the Motor Vehicle Collisions data. Variable "Year" indicates the year in which the accident occurs. Variables "Street1" and "Street2" provides the name of the street where the accident takes place. Variable "District" indicates its district. Variable "Collision location" includes 8 types: at intersection, intersection related, at/near private drive, laneway, non intersection, overpass or bridge, trail, and underpass or tunnel. Variable "Classification of Accident" involves fatal and non-fatal injury.

@@

```{r fig.cap = "Comparing the number of traffic collisions each year from 2006 to 2020", fig.width=5.4,fig.height=3.6, echo=FALSE}
ggplot(mapping = aes(x = year, fill = acclass), data=vehicle_collision) +
  geom_bar()+theme_light()+ylab("number of motor vehicle collisions involving injuries or deaths")+scale_fill_discrete(name = "Classification of Accident")
```
Figure 1 shows a descending pattern for the number of car accidents involving injuries or deaths as the number went from 1483 cases in 2006 to 633 cases in 2020. By looking at the accident types,  Indicate that it's likely that Toronto has done a better job on traffic control in general. However, if other factors are taken into considerations, the trend of motor

```{r fig.cap = "Comparing the location of traffic collisions each year from 2006 to 2020"}
ggplot(mapping = aes(x = year, fill=accloc), data=vehicle_collision) +
  geom_bar()
```

```{r bills, fig.cap="Bills of penguins", echo = FALSE}
ggplot(penguins, aes(x = island, fill = species)) +
  geom_bar(alpha = 0.8) +
  scale_fill_manual(values = c("darkorange","purple","cyan4"),
                    guide = FALSE) +
  theme_minimal() +
  facet_wrap(~species, ncol = 1) +
  coord_flip()
```

Talk more about it.

Also bills and their average (Figure \@ref(fig:billssssss)). (Notice how you can change the height and width so they don't take the whole page?)

```{r billssssss, fig.cap="More bills of penguins", echo = FALSE, fig.width=8, fig.height=4}

# This needs to be about the random data tha tI sasmpled?
ggplot(penguins, aes(x = island, fill = species)) +
  geom_bar(alpha = 0.8) +
  scale_fill_manual(values = c("darkorange","purple","cyan4"),
                    guide = FALSE) +
  theme_minimal() +
  facet_wrap(~species, ncol = 1) +
  coord_flip()
```

Talk way more about it.

# Model

\begin{equation}
Pr(\theta | y) = \frac{Pr(y | \theta) Pr(\theta)}{Pr(y)}  (\#eq:bayes)
\end{equation}

Equation \@ref(eq:bayes) seems useful, eh?

Here's a dumb example of how to use some references: In paper we run our analysis in `R` [@citeR]. We also use the `tidyverse` which was written by @thereferencecanbewhatever If we were interested in baseball data then @citeLahman could be useful.

We can use maths by including latex between dollar signs, for instance $\theta$.

# Results

@@

>>>>

---
title: "Investigating motor vehicle collisions in Toronto: Installing more red light cameras do not stop traffic accidents from happening"
author: 
  - Alicia Yang
thanks: "Code and data are available at: https://github.com/Alicia-y/Telling-stories-with-data."
date: "`r format(Sys.time(), '%d %B %Y')`"
abstract: "Vehicle collisions occur constantly in big citys. They sometimes could be severe issues when resulting in damage in public and personal properties, injuries, or even death. This report aims to investigate motor vehicle collisions in Toronto that result in injuries or deaths from 2006 to 2020. By exploring the trend of these traffic collisions in respect of their frequency, fatality, and location, I find that the total number of accidents has a descending trend. However, the number of vehicle collisions that occurred at intersections has not reduced despite the fact that more red light cameras are installed. This finding suggests that ... furthur measures should be taken to enhance the safety issue on road."
output:
  bookdown::pdf_document2
toc: TRUE
bibliography: references.bib
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
# Set up libraries
library(tidyverse)
library(janitor)
```

@@

As one of the busiest and most modern city in Canada, Toronto constantly faces a large number of citizens, and, inevitably, heavy traffic. The road condition in Toronto is busy most of the time. The vast majority of roads and buildings under construction that take place almost everywhere narrow the roadway of vehicles and turn off traffic lights. Heavy traffic is likely to be one of the main reasons behind car accidents, because it increases the road rage and encourages the vehicle drivers to make dangerous moves to save time. Vehicle collisions have always brought large financial costs to the involving parties, and even more, they even brought injuries and deaths. In just 2019, traffic accidents resulted in more than 1700 deaths and more than 140000 injuries.

The Motor Vehicle Collisions data of Toronto provides a record of the past severe vehicle collisions in Toronto which result in injuries or deaths, and with the help of statistical data analysis, it can possibly offer insights on the ways to reduce this type of accident. This report investigates the motor vehicle collisions in Toronto from 2006 to 2020, with a focus on the ones result in injuries or deaths. In Section \@ref(data), the Motor Vehicle Collisions dataset is introduced and organized to visualize the change in the annual number of traffic accidents with respect to their fatality and their location. Then, I discussed about the possible implications of these visualizations. By Figure 2, I concluded that motor vehicle collisions are always more likely to occur at intersections. Thus, 
Continuing

@@

# Data
## Motor Vehicle Collisions involving Killed or Seriously Injured Persons
To investigate the motor vehicle collisions in Toronto, I obtained the dataset "Motor Vehicle Collisions involving Killed or Seriously Injured Persons" from the Toronto Open Data Portal [@citedata]. Provided by the Toronto Police Service, this dataset includes all vehicle collisions occurred in Toronto between 2006 and 2020 that result in injuries or deaths. One bias aroused by this dataset is the offset from the real accident location. For ethical reason and the protection of the parties involved in the accidents that are considered to be criminal cases, the location of those criminal accidents are offset to the closest intersection node [@citecollisiondata]. As a result, the location of those accidents specific to neighborhoods might not be accurate. Therefore, I will use the variable district instead of neighbourhood to categorize the accidents later in the report to avoid this bias.

This dataset contains 16840 observations of collisions and 54 variables with all kinds of information of each collision. This report focuses on four of these variables which are: year, accident location, district, and fatality. By using R packages "tidyverse" [@citetidyverse], "janitor" [@citejanitor], "kableExtra" [@citekableExtra] and "knitr" [@citeknitr], an extract of the cleaned dataset (Table \@ref(tab:dataextract)) is shown below.

```{r, echo=FALSE, message=FALSE, warning=FALSE}
# Load motor vehicle collision data and clean
vehicle_collision <- read_csv(here::here("inputs/data/vehicle_collisions.csv")) %>%
  clean_names() %>%
  drop_na(year) %>%
  filter(acclass != "Property Damage Only")
```

```{r dataextract, echo=FALSE, message=FALSE}
# Set up libraries
library(knitr)
library(kableExtra)

# Create a table with selected columns
vehicle_collision <- vehicle_collision %>%
  select(year, accloc, district, acclass)
vehicle_collision %>%
  slice(1:10) %>%
  kable(
    caption = "Extracting the first ten rows from the Motor Vehicle Collisions data",
col.names=c("Year", "Collision location", "District", "Fatality"),
linesep = "", digits = 1, booktabs=TRUE) %>%
  kable_styling(latex_options = "HOLD_position", font_size = 10)

@@

Table \@ref(tab:dataextract) shows the first ten rows of the Motor Vehicle Collisions data. Variable "Year" indicates the year in which the accident occurs. Variable "District" provides the district of the accident. Variable "Collision location" includes 8 types: at intersection, intersection related, at/near private drive, laneway, non intersection, overpass or bridge, trail, and underpass or tunnel. Variable "Fatality" involves fatal and non-fatal injury.

@@

```{r fig.cap = "Comparing the number of traffic collisions each year from 2006 to 2020", fig.width=8.4,fig.height=3.6, echo=FALSE}
# Create a bar plot to show number of collisions by fatality
ggplot(mapping = aes(x = year, fill = acclass), data=vehicle_collision) +
  geom_bar()+theme_light()+ylab("number of motor vehicle collisions")+scale_fill_discrete(name = "Fatality")+facet_wrap(~acclass)
```

Figure 1 shows a descending pattern for the number of car accidents involving non-fatal injuries from year 2006 to 2020. The number of fatal accidents does not increase nor decrease much. In general, the number vehicle collisions in Toronto has largely decrease, indicating that when combining this two factors(number and fatality), Toronto has done a better job on traffic control in general. I'm also interested in the trend of motor vehicle collisions if other factors are taken into considerations. 
\newpage

```{r fig.cap = "Comparing the location of traffic collisions each year from 2006 to 2020", echo=FALSE, fig.width=6,fig.height=3.2}
# Create a bar plot to show number of collisions by accident location
ggplot(mapping = aes(x = year, fill=accloc), data=vehicle_collision)+scale_fill_discrete(name = "Accident location")+ylab("number of motor vehicle collisions")+theme_light()+geom_bar()
```

Figure 2 illustrates the where these vehicle collisions usually take place from 2006 to 2020. Before 2013, half of the data doesn't come with accident location, which is understandable because back in the days there wasn't an easy approach to record and save the data, so they often got lost. A large proportion of the accidents take place either at intersections or is intersection related. The other seven accident locations types add up to less than half of all vehicle collisions. At intersections, a lot more rule breaking behaviors can be committed by drivers. For instance, they could sometimes ignore stop signs, speed up when approaching yellow traffic lights; some even claimed that they rarely yielded at intersections. These behaviors largely increase the possibility of them colliding with other vehicles and the pedestrians at intersections.

Motor vehicle collisions are most likely to take place at intersections. Thus, if measures are taken to restrict the driver's rule breaking behaviors at intersections, the numbers of car accidents would certainly drop a lot in general. Red light camera is claimed to be a measure taken by the government to monitor the vehicles at intersections. I will look into whether it's effective or not in respect of preventing the traffic accidents from happening.

## Red Light Cameras
The number of red light camera instilled in Toronto from 2007 to 2020 is obtained from the dataset "Red Light Cameras" found in the Toronto Open Data Portal [@citedata]. This dataset is shared by the Transportation Services and it covers the locations of red light cameras in Toronto. In its introduction, this dataset claims that red light cameras can improve the safety of the drivers and pedestrians by monitoring the behavior of the drivers [@citecameradata]. It contains 199 observations and 32 variables. This dataset's downside is that it is missing values in lots of its variables. To reduce errors generated from this downside, I will only utilize the variables with no missing values. R packages "tidyverse" [@citetidyverse], "janitor" [@citejanitor], "kableExtra" [@citekableExtra] and "knitr" [@citeknitr] are usedt to produce an extract of the cleaned dataset (Table \@ref(tab:dadaextract1)).

```{r dadaextract1, message=FALSE, warning=FALSE, echo=FALSE}
# Load red light camera data and clean
red_light_camera <- read_csv(here::here("inputs/data/red_light_camera.csv")) %>%
  clean_names() %>%
  select(activation_date, district) %>%
  drop_na(activation_date) %>%
  drop_na(district)

# Create "year" variable
red_light_camera$year <- vector("numeric", length = 199)
for (i in 1:199){
  red_light_camera$year[i]=format(as.Date(red_light_camera$activation_date[i], format="%Y/%m/%d"),"%Y")
}

# Create a table with selected columns
red_light_camera %>%
  select(district, year, activation_date) %>%
  slice(1:10) %>%
  kable(
    caption = "Extracting the first ten rows from the Red Light Camera data",
col.names=c("District", "Year", "Activation date"),
linesep = "", digits = 1, booktabs=TRUE) %>%
  kable_styling(latex_options = "HOLD_position", font_size = 10)
```

Table \@ref(tab:dadaextract1) shows the first ten rows of the Red Light Camera data in Toronto. "Year" is a new variable added by me into this dataset. It is pulled from the original variable "activation date", and it provides the year in which the red light camera is activated. Variable "District" includes the four main districts of Toronto: Toronto and East York, North York, Scarborough, and Etobicoke York.

I'm interested in observing the cumulative number of red light cameras installed in each district from 2006 to 2020 and compare this data with the trend of number of motor vehicle collisions to see if they actually have connections.

```{r message=FALSE, warning=FALSE, echo=FALSE, fig.cap = "Number of red light cameras in the four districts each year", fig.width=7,fig.height=3.2}
# Create a new dataset that presents the total number of red light cameras in each district
red_light_camera <- red_light_camera %>%
  filter(year != 2021)
cumulative <-tibble(
  year=c(2006, 2007, 2008, 2009, 2010, 2011, 2012, 2013, 2014, 2015, 2016, 2017, 2018, 2019, 2020, 2006, 2007, 2008, 2009, 2010, 2011, 2012, 2013, 2014, 2015, 2016, 2017, 2018, 2019, 2020, 2006, 2007, 2008, 2009, 2010, 2011, 2012, 2013, 2014, 2015, 2016, 2017, 2018, 2019, 2020, 2006, 2007, 2008, 2009, 2010, 2011, 2012, 2013, 2014, 2015, 2016, 2017, 2018, 2019, 2020),
  sum=c(0, 2, 6, 15, 16, 17, 17, 17, 17, 17, 17, 31, 33, 33, 34, 0, 0, 7, 17, 18, 18, 18, 18, 18, 18, 18, 42, 43, 44, 44, 0, 3, 10, 18, 19, 19, 19, 19, 19, 19, 19, 26, 27, 28, 28, 0, 1, 9, 15, 15, 15, 15, 15, 15, 15, 15, 36, 38, 41, 41),
  district=c("Toronto and East York", "Toronto and East York", "Toronto and East York", "Toronto and East York", "Toronto and East York", "Toronto and East York", "Toronto and East York", "Toronto and East York", "Toronto and East York", "Toronto and East York", "Toronto and East York", "Toronto and East York", "Toronto and East York", "Toronto and East York", "Toronto and East York", "Scarborough", "Scarborough", "Scarborough", "Scarborough", "Scarborough", "Scarborough", "Scarborough", "Scarborough", "Scarborough", "Scarborough", "Scarborough", "Scarborough", "Scarborough", "Scarborough", "Scarborough", "North York", "North York", "North York", "North York", "North York", "North York", "North York", "North York", "North York", "North York", "North York", "North York", "North York", "North York", "North York", "Etobicoke York", "Etobicoke York", "Etobicoke York", "Etobicoke York", "Etobicoke York", "Etobicoke York", "Etobicoke York", "Etobicoke York", "Etobicoke York", "Etobicoke York", "Etobicoke York", "Etobicoke York", "Etobicoke York", "Etobicoke York", "Etobicoke York")
)

#Create a line plot from the dataset
ggplot( aes(x=year, y=sum, group=district, color=district), data=cumulative) +
    geom_line()+ylab("number of red light cameras")+theme_light()
```

Figure 3 demonstrates the total number of red light cameras in each district from 2006 to 2020. Scarborough installed the most red light cameras in this period of time, especially after 2016, a large number of red light cameras were installed there. The district of North York, on the contrary, started with the most red light cameras installed among all districts, but has the slowest growth and ended up with the least number of red light cameras in 2020 among all districts.

I wanted to see how the increase in number of red light cameras compare with the traffic collision's frequency in each district. So, I selected the accidents that occurred at intersections or somewhere near intersections from the Motor Vehicle Collision dataset and created a ggplot to see their trend by each district in Toronto.

```{r, echo=FALSE, warning=FALSE, fig.cap = "Comparing the number of motor vehicle collisions take place at or near intersections by district", fig.width=7,fig.height=3.2}
# Select the accidents that only occur at or near intersections
co <- vehicle_collision %>%
  filter(accloc == "At Intersection" | accloc == "Intersection Related") %>%
  drop_na(district)
for(i in 1:9039){
  if(co$district[i] == "Toronto East York"){
    co$district[i] = "Toronto and East York"
  }
}

# Create bar plot to compare vehicle collisions by districts
ggplot(data=co,aes(x=year, fill=acclass))+geom_bar()+facet_wrap(~district)+ylab("number of motor vehicle collisions")+theme_light()+scale_fill_discrete(name = "Fatality")
```

Figure 4 shows the number of annual motor vehicle accidents occurred at or near intersections in each district. For all the districts, the numbers of fatal vehicle collisions stay approximately the same throughout these years, with Scarborough having the highest mortality rate in general. With regards to the total number of accidents, data in Etobicoke York, North York, and Scarborough all has its fluctuations, but the number remains about the same. The number of vehicle collisions in Toronto and East York was the highest in 2006, with more than to 260 cases, but it is the only district where the number has dropped.

By comparing Figure 3 and Figure 4, I found that they haven't show any associations between the number of traffic accidents and the number of red light cameras installed.

@@

https://www.scientificamerican.com/article/red-light-cameras-may-not-make-streets-safer/
